Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-07-11 submission/Response_to_reviewers.md

**Reviewer #1 (Comments for the Author):

Major Comments
• P.5 (line 81): The bioinformatics behind the statistical analyses are important. Include a brief summary in the main text. Specifically, it is important to clarify that you focused on studies that collected targeted 16S sequence data (not shotgun metagenomes). Also briefly mention that you did OTU clustering from scratch (per study and pooled?). Here or in the methods, clarify how alpha and beta diversity statistics were estimated. **

**• P.7 (lines 134-136): To overcome challenges with non-overlapping 16S regions, might it be possible to predict taxonomy of OTUs (which it appears you have already done for phylum level analyses) and then do a pooled community structure analysis on the taxa rather than OTUs? Alternatively, perhaps tools designed for identifying OTUs from non-overlapping shotgun marker gene reads might work. **

**• P.8 (line 162): What variability was used in the calculations (presumably that observed in the studies)? For designing future studies, it would be great to investigate the sensitivity of the results to the amount of between-sample variance, since variance does differ between studies (e.g., is higher in rural populations and is lower between moms and their newborns than between unrelated individuals). **

**• P.10 (line 204): When discussing biological relevance, it would be useful to also mention that relative abundance may not be the relevant factor for the host. Imagine an OTU (A) that produces a metabolite with a physiological effect on obesity (or whatever host trait one is studying). Suppose the number of cells and the amount of metabolite produced remains constant, but the number of cells for some other OTU (B) goes up by an order of magnitude. The relative abundance of A in the community has gone down, but this may be irrelevant to the molecular/cellular mechanism affecting obesity, especially if B has no relationship to A or obesity (e.g., it occupies a distinct niche in the gut). The same argument probably holds for alpha diversity metrics based on relative abundances - the host likely cares about specific components of the diversity not diversity per se. **

**• P.12 (line 244): Where can the pipeline be obtained? This info is in the methods but should probably be highlighted.**

**• P.12 (lines 265-267): Elaborate further on this point. The right biomarker may be a protein or pathway. Even if this is taxonomically restricted, it may be hard to detect with community-level statistics or even OTU relative abundances. In addition to the issue noted above regarding relative abundance, species-level OTUs can still miss functional variability due to large differences in gene content among strains with identical or nearly identical 16S sequences. **

**Minor Comments
• P.4 (lines 53-56): repeated text and typos: "mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities were manipulated with antibiotics or underwent colonization with varied communities appears to support the association since these manipulation yielded" **

**P.6 (line 100) and elsewhere: When first referring to a study by name (e.g., Walters, Finucane, Goodrich) provide the citation. This info is in Table 1, but it would be helpful to provide in the text as well. **

**P.6 (line 103): In the random effects model does study have a random intercept or slope (or both)? **

**P.6 (line 88): What is the overlap of studies with the Walters and Finucane analyses (could move or copy this from the discussion)? **

**P.6 (line 88): What is the range of sample size per study (in Table 1, but would be helpful to give this overview statistic in the text)? **

**P.7 (lines 122-123): Clarify if ranges on relative risk are 95% confidence intervals or some other statistic. **

**P.7 (line 123): Fix plural: "were no significant difference" **

**P.7 (lines 132-133): Explain why directionality cannot be inferred. **

**P.13 (line 281): Is this algorithm the same as agglomerative hierarchical clustering with average linkage? Also, provide some more detail on how the read counts for OTUs were used to estimate the various alpha and beta diversity statistics. **

**P.14 (line 297): Typo: "each study within a study" **

**Reviewer #2 (Comments for the Author):

General comments:
Overall this paper is a useful meta analysis of obesity and the microbiome. I have some miscellaneous comments that are below. In addition I have one more important concern that I believe needs to be addressed.

More important concern:
Given that their analysis depends on them compiling data from multiple studies and reanalyzing that data, I believe it is necessary for these authors to make sure that all components of this analysis are reproducible. I commend the authors for their commitment to sharing code for all the work. However, I have a few concerns about the data. First, some of the data is from a personal web site. I would recommend that the authors here either repost that data to a more sustainable site (e.g., Figshare), get the original producers of the data to post it to a more sustainable site, or come up with some other solution. Again, I commend the authors for their efforts in openness and reproducibility, I am just concerned that some aspects of this paper will not be reproducible. **

**Additional concerns and comments:
L47 "individuals had a lower diversity than lean individuals (6)." Specify what kind of diversity **

**L54 "with antibiotics or colonization with varied communities were manipulated with antibiotics" Typo? Should this be "that were manipulated?" **

**L73 "Literature Review and Study Inclusion. We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis (14)." Would be good to say something about whether this is considered an important approach in meta-analysis **

**L77 "we searched PubMed for original research studies that involved studying obesity and the..." Given the importance of covering the literature well in meta-analyses it would be good to say how Pubmed was searched.**

**L78 "We identified ten additional studies" How?**

**L80 "and obesity. We then manually curated the 196 studies to select." Numbers don't add up - 187 + 10 = 197. I looked through the Figure and I am not sure this explains all the numbers either.**

**L102 "other studies appeared to have the same trend, albeit the differences were not statistically significant." Quibble about wording here. If the result is not significant - is it a trend?**

**L110 "significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant." Some justification for this statement is needed.**

**L126 "obese, it is questionable whether that risk is biologically or clinically relevant." Again, Some justification for this statement is needed.**

**L164 "each of the studies (Figures 6, S3-S8). Although there is no biological rationale for these effect sizes, they represent a range that is plausible." It would help to clarify what is meant here by "range that is plausible"?**

**L193 "datasets. This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome was limited" I think it would be good here to clarify that this is really the ability to classify based solely on microbiome data. It might be possible to do this much better if individuals were subclassified (e.g., by genotype, age, gender, etc)

**Some references could / should be added: lme4 (v.1.1-12) R package, vegan (v.2.3-5) R package, AUCRF (1.1) R package, pROC (1.8) R package, pwr (1.1-3) R package**

**Table 1 - would be good to add the references for the studies**

**Figure 1 - some of the boxes in the flow chart are cutoff.**

### 2016-07-11 submission/Sze_Obesity_mBio_2016.Rmd
Respond to reviewer's comments and edit text accordingly

There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans. The first such study used genetically obese mice and observed the ratio of the relative abundances of *Bacteroidetes* to *Firmicutes* (B:F) was lower in obese mice than lean mice [@ley_obesity_2005]. Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009]. They also showed that the relative abundance of *Bacteroidetes* and *Firmicutes* increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006]. Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities
were manipulated with antibiotics or underwent colonization with varied  communities appears to support the association since these manipulation yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016]. The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.

The recent publication of additional studies that collected BMI data for each subject as well as other studies that were not included in the earlier re-analysis studies offered the opportunity to revisit the question relating the structure of the human microbiome to obesity. One critique of the prior re-analysis studies is that the authors did not aggregate the results across studies to increase the effective sample size. It is possible that there were small associations within each study that were not statistically significant because the individual studies lacked sufficient power. Alternatively, diversity metrics may mask the appropriate signal and it is necessary to measure the association at the level of microbial populations. Walters et al. [-@walters_meta-analyses_2014] demonstrated that Random Forest machine learning models were capable of predicting obesity status within a single cohort, but did not attempt to test the models on other cohorts. The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously.

@@

***Literature Review and Study Inclusion.***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplemental Text and in Figure 1. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified `r number_string[n]` additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 10 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The `r number_string[n]` studies are summarized in Table 1.

@@

	paste0(format(data$rr, digits=2, nsmall=2L), " (", format(data$ci_lb, digits=2, nsmall=2L), "-", format(data$ci_ub, digits=2, nsmall=2L), ")")

@@

***Relative risk.***
Building upon the alpha diversity analysis we calculated the relative risk of being obese based on an individual's alpha diversity metrics relative to the median metric for that study. Inspection of funnel plots for each of the metrics suggested that the studies included in our analysis were not biased (Figure S2). The results using relative risk largely matched those of using the raw alpha diversity data. Across the `r number_string[n]` studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S3). Again, although the relative risk values were not significant for other studies, the values tended to be above one. When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`). The relative risks associated with alpha diversity were small. The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`. There were no significant difference in the phylum-level metrics. Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.

@@

***Beta diversity analysis.***
Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study. We then used AMOVA to test for significant differences between the structure of non-obese and obese individuals (Table 1). The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05). Because it was not possible to ascertain the directionality of the difference in community structure nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.

@@

***Power and Sample Size Estimate Simulations.***
The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals. To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S4-S9). Although there is no biological rationale for these effect sizes, they represent a range that is plausible. Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6). None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S4). In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`. Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15. We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups. To detect a 1, 5, 10, or 15% difference in Shannon index, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in B:F values, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median required sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric. In the case of obesity, the studies were underpowered to detect the `r signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.

@@

Our meta-analysis helps to provide clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity. We performed an extensive literature review of the existing studies on the microbiome and obesity and performed a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. By statistically pooling the data from ten studies, we observed significant, but small, relationships between richness, evenness, and diversity and obesity status as well as the relative risk of being obese based on these metrics. We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets. This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome was limited. Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes. Considering these datasets are among the largest published, it appears that most human microbiome studies are underpowered to detect differences in alpha diversity.

@@

By selecting a range of possible effect sizes, we were able to demonstrate that most studies are underpowered to detect modest differences in alpha diversity metrics and phylum-level relative abundances. Several factors interact to limit the power of microbiome studies. There is wide interpersonal variation in the diversity and structure of the human microbiome. In addition, the common experimental designs limit their power. As we observed, most of the studies included in our analysis were unbalanced for the variable that we were interested in. This was also true of those studies that originally sought to identify associations with obesity. Even with a balanced design, we showed that it was necessary to obtain approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)` and `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)` samples per group to detect a 5% difference in Shannon diversity or B:F, respectively. It was interesting that these sample sizes agreed across studies regardless of their sequencing method, region within the 16S rRNA gene, or subject population (Figure 6). This suggests that regardless of the treatment or category, these sample sizes represent a good starting point for subject recruitment when using stool samples. Unfortunately, few studies have been published with this level of subject recruitment. This is troubling since the positive predictive rate of a significant finding in an underpowered study is small leading to results that cannot be reproduced [@Ioannidis2005]. Future microbiome studies should articulate the basis for their experimental design.

Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference. Additionally, our analysis used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012]. Our analysis included four of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded three of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012]. The additional `r number_string[n-3]` datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies. Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study. To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available. Similarly, it would be possible to adapt this pipeline to other body sites and treatment or variables (e.g. subject's sex or age).

Similar to our study, the Walters et al [@walters_meta-analyses_2014] generated Random Forest machine learning models to differentiate between non-obese and obese individuals. They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis. When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets. This could be due to differences in subject populations and methods. Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.

Although we failed to find an effect this does not necessarily mean that there is no role for the microbiome in obesity. There is strong evidence in murine models of obesity that the microbiome and level of adiposity can be manipulated via genetic manipulation of the animal and manipulation of the community through antibiotics or colonizing germ free mice with diverse fecal material from human donors [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016; @ley_obesity_2005]. These studies appear to conflict with the observations using human subjects. Recalling the large interpersonal variation in the structure of the microbiome, it is possible that each individual has their own signatures of obesity. Alternatively, it could be that the involvement of the microbiome in obesity is at the level of a common set of metabolites that can be produced from different structures of the microbiome.

>>>>

There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans. The first such study used genetically obese mice and observed the ratio of the relative abundances of *Bacteroidetes* to *Firmicutes* (B:F) was lower in obese mice than lean mice [@ley_obesity_2005]. Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower alpha-diversity than lean individuals [@turnbaugh_core_2009]. They also showed that the relative abundance of *Bacteroidetes* and *Firmicutes* increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006]. Two re-analysis studies by Walters et al. [@walters_meta-analyses_2014] and Finucane et al. [@finucane_taxonomic_2014] interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized. Regardless of the results using human populations, studies using animal models where the community was manipulated with antibiotics or established by colonizing germ-free animals  with varied communities appear to support the association since these manipulations yielded differences in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016]. The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.

The recent publication of additional studies that collected BMI data for each subject as well as other studies that were not included in the earlier re-analysis studies offered the opportunity to revisit the question relating the structure of the human microbiome to obesity. One critique of the prior re-analysis studies is that the authors did not aggregate the results across studies to increase the effective sample size. It is possible that there were small associations within each study that were not statistically significant because the individual studies lacked sufficient power. Alternatively, diversity metrics may mask the appropriate signal and it is necessary to measure the association at the level of microbial populations. The Walters re-analysis study demonstrated that Random Forest machine learning models were capable of predicting obesity status within a single cohort, but did not attempt to test the models on other cohorts. The purpose of this study was to perform a meta-analysis of the association between differences in the microbiome and obesity status by analyzing and applying a more systematic and synthetic approach than was used previously.

@@

***Literature Review and Study Inclusion.***
To perform a robust meta-analysis and limit inclusion bias, we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify the studies that we analyzed [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplemental Text and in Figure 1. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified `r number_string[n]` additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and 16S rRNA gene sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 10 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The `r number_string[n]` studies are summarized in Table 1. For comparison, two of these studies were included in the Finucane re-analysis study [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012] and four of these studies were included in the Walters re-analysis study [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011]. The 16S rRNA gene sequence data from each study was re-analyzed using a similar approach based on previously described methods for reducing the number of chimeric sequences and sequencing errors for 454 and Illumina MiSeq data [@Kozich_2013; @Schloss2011]. The sequences were either clustered into operational taxonomic units (OTUs) using the average neighbor approach [@Westcott2015] or taxonomic groupings based on their classification using a naive Bayesian classifier [@Wang2007].

@@

	paste0(format(data$rr, digits=2, nsmall=2L), " (95% CI: ", format(data$ci_lb, digits=2, nsmall=2L), "-", format(data$ci_ub, digits=2, nsmall=2L), ")")

@@

***Relative risk.***
Building upon the alpha diversity analysis we calculated the relative risk of being obese based on an individual's alpha diversity metrics relative to the median metric for that study. Inspection of funnel plots for each of the metrics suggested that the studies included in our analysis were not biased (Figure S2). The results using relative risk largely matched those of using the raw alpha diversity data. Across the `r number_string[n]` studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S3). Again, although the relative risk values were not significant for other studies, the values tended to be above one. When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`). The relative risks associated with alpha diversity were small. The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`. There were no significant differences in the phylum-level metrics. Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.

@@

***Beta diversity analysis.***
Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study. We then used AMOVA to test for significant differences between the structure of non-obese and obese individuals (Table 1). The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05). Because it was not possible to ascertain the directionality of the difference in community structure because the samples are arrayed in a non-dimensional space. nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.

@@

***Power and Sample Size Estimate Simulations.***
The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals. To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S4-S9). Although there is no biological rationale for these effect sizes, they represent a range that includes effect sizes that would be generally considered to be biologically significant. Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6). None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S4). In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`. Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15. We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups. To detect a 1, 5, 10, or 15% difference in Shannon index, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in B:F values, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median required sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric. In the case of obesity, the studies were underpowered to detect the `r signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.

@@

Our meta-analysis helps to provide clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity. We performed an extensive literature review of the existing studies on the microbiome and obesity and performed a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. By statistically pooling the data from ten studies, we observed significant, but small, relationships between richness, evenness, and diversity and obesity status as well as the relative risk of being obese based on these metrics. We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets. This analysis demonstrated that the ability to reliably classify individuals as being obese based solely on the composition of their microbiome was limited. Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes. Considering these datasets are among the largest published, it appears that most human microbiome studies are underpowered to detect differences in alpha diversity.

@@

By selecting a range of possible effect sizes, we were able to demonstrate that most studies are underpowered to detect modest differences in alpha diversity metrics and phylum-level relative abundances. Several factors interact to limit the power of microbiome studies. There is wide interpersonal variation in the diversity and structure of the human microbiome. Some factors such as relationship between subjects could potentially decrease the amount variation [@turnbaugh_core_2009] and other factors such as whether one lives in a rural environment could increase the amount of variation [@yatsunenko_human_2012]. It was interesting that although there were ranges in variation across the studies included in our analysis, the sample size calculations yielded similar results. In addition, the common experimental designs limit their power. As we observed, most of the studies included in our analysis were unbalanced for the variable that we were interested in. This was also true of those studies that originally sought to identify associations with obesity. Even with a balanced design, we showed that it was necessary to obtain approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)` and `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)` samples per group to detect a 5% difference in Shannon diversity or B:F, respectively. It was interesting that these sample sizes agreed across studies regardless of their sequencing method, region within the 16S rRNA gene, or subject population (Figure 6). This suggests that regardless of the treatment or category, these sample sizes represent a good starting point for subject recruitment when using stool samples. Unfortunately, few studies have been published with this level of subject recruitment. This is troubling since the positive predictive rate of a significant finding in an underpowered study is small leading to results that cannot be reproduced [@Ioannidis2005]. Future microbiome studies should articulate the basis for their experimental design.

Two previous re-analysis studies have stated that there was not a consistent association between alpha diversity and obesity [@walters_meta-analyses_2014; @finucane_taxonomic_2014]; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference. Additionally, our analysis used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012]. Our analysis included four of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded three of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012]. The additional `r number_string[n-3]` datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies. Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study. To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available (https://github.com/SchlossLab/Sze_Obesity_mBio_2016/). Similarly, it would be possible to adapt this pipeline to other body sites and treatment or variables (e.g. subject's sex or age).

Similar to our study, the Walters study generated Random Forest machine learning models to differentiate between non-obese and obese individuals [@walters_meta-analyses_2014]. They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis. When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets. This could be due to differences in subject populations and methods. Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.

Although we failed to find an effect this does not necessarily mean that there is no role for the microbiome in obesity. There is strong evidence in murine models of obesity that the microbiome and level of adiposity can be manipulated via genetic manipulation of the animal and manipulation of the community through antibiotics or colonizing germ free mice with diverse fecal material from human donors [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016; @ley_obesity_2005]. These studies appear to conflict with the observations using human subjects. Recalling the large interpersonal variation in the structure of the microbiome, it is possible that each individual has their own signatures of obesity. Alternatively, it could be that the involvement of the microbiome in obesity is not apparent based on the taxonomic information provided by 16S rRNA gene sequence data. Rather, the differences could become more apparent at the level of a common set of gene transcripts or metabolites that can be produced from different structures of the microbiome.
